---
title: "School data in the LMA"
author: "Joey Talbot"
format: gfm
prefer-html: true
---

```{r setup, echo = FALSE}
library(tidyverse)
library(sf)
library(tmap)
tmap_mode("view")
```

## School locations and number of students

We used school location data from [Carris Metropolitana](https://raw.githubusercontent.com/carrismetropolitana/datasets/latest/facilities/schools/schools.csv), plus national data on school type and enrolled student numbers by gender and year group from [the General Directorate of Education and Science Statistics](https://www.dgeec.medu.pt/api/ficheiros/65708d9fcc461a4d46cd9a13).

We sourced high resolution population data, including the population aged 0-14 living within small zones, from [BGRI](). 



```{r}
schools_geo = readRDS("data/SCHOOLS_geo.Rds")
schools_year = readRDS("data/SCHOOLS_year.Rds")

bgri = read_sf("../internal/BGRI21_170/BGRI21_170.gpkg")
bgri_lisbon = bgri |> 
  filter(DTMN21 == 1106)

schools_both = inner_join(schools_year, schools_geo, by = c("DGEEC_id" = "CÓDIGO DGEEC ESCOLA"))
schools_both = st_as_sf(schools_both)

by_ciclo = schools_both |> 
  group_by(DGEEC_id, ESCOLA, MUNICIPIO, PUBLIC_PRIVATE, CICLO_ESTUDOS) |> 
  summarise(STUDENTS = sum(STUDENTS))

c1 = by_ciclo |> 
  filter(CICLO_ESTUDOS == "1.º Ciclo")
c2 = by_ciclo |> 
  filter(CICLO_ESTUDOS == "2.º Ciclo")
c3 = by_ciclo |> 
  filter(CICLO_ESTUDOS == "3.º Ciclo")

# saveRDS(c1, "data/c1.Rds")
# saveRDS(c2, "data/c2.Rds")
# saveRDS(c3, "data/c3.Rds")

c1_public = by_ciclo |> 
  filter(CICLO_ESTUDOS == "1.º Ciclo",
         PUBLIC_PRIVATE == "Público")
c2_public = by_ciclo |> 
  filter(CICLO_ESTUDOS == "2.º Ciclo",
         PUBLIC_PRIVATE == "Público")
c3_public = by_ciclo |> 
  filter(CICLO_ESTUDOS == "3.º Ciclo",
         PUBLIC_PRIVATE == "Público")

# saveRDS(c1_public, "data/c1_public.Rds")
# saveRDS(c2_public, "data/c2_public.Rds")
# saveRDS(c3_public, "data/c3_public.Rds")
```


```{r}
lisbon_c1 = by_ciclo |> 
  filter(MUNICIPIO == "Lisboa",
         CICLO_ESTUDOS == "1.º Ciclo")
lisbon_c2 = by_ciclo |> 
  filter(MUNICIPIO == "Lisboa",
         CICLO_ESTUDOS == "2.º Ciclo")
lisbon_c3 = by_ciclo |> 
  filter(MUNICIPIO == "Lisboa",
         CICLO_ESTUDOS == "3.º Ciclo")

tm_shape(bgri_lisbon) + tm_polygons("N_INDIVIDUOS_0_14") +
  tm_shape(lisbon_c1) + tm_dots(size = "STUDENTS")
```

```{r}

```

```{r}

```