---
title: "From Origin-Destination Data to Bike Bus Routes"
vignette: >
  %\VignetteIndexEntry{fromOD2bikebus}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| eval: false
#| include: false

devtools::load_all()
```

This vignette demonstrates the complete workflow for generating bike bus routes from origin-destination data using the bicischools package. We'll walk through the process using data from Almada, Portugal, showing how to go from raw geographic data to optimized bike bus routes.

## Setup

```{r}
#| label: setup
#| message: false

# library(bicischools)
library(sf)
library(tidyverse)
library(tmap)
tmap_mode("view")
```

## Overview

The workflow consists of several key steps:

1. **Data Preparation**: Load and process geographic zones and school locations
2. **Origin-Destination Simulation**: Use gravity models to estimate student trips
3. **Route Generation**: Create cycling routes between origins and destinations  
4. **Network Analysis**: Build route networks and calculate cycling uptake
5. **Route Optimization**: Identify and optimize bike bus routes
6. **Visualization**: Map results for analysis

## 1. Data Preparation

We'll use census zones (BGRI) as origins and schools as destinations. First, let's load the included sample data:

```{r}
#| label: load-data
# Load sample data included with the package
data("origins_almada")
data("schools_lisbon")
data("od_data_almada")
data("routes_almada")

# Display basic information
cat("Number of origin zones:", nrow(origins_almada), "\n")
cat("Number of schools:", nrow(schools_lisbon), "\n")
cat("Number of OD pairs:", nrow(od_data_almada), "\n")
```

Let's visualize the basic geography:

```{r}
#| label: basic-map
tm_shape(origins_almada) +
    tm_borders(alpha = 0.5) +
    tm_shape(schools_lisbon) +
    tm_dots(col = "red", size = 0.1) +
    tm_layout(title = "Origins and Schools in Almada")
```

## 2. Origin-Destination Flow Simulation

For this example, we'll use pre-calculated OD data, but here's how you would generate it using a gravity model:

```{r}
#| label: gravity-model
#| eval: false

# Example gravity model simulation (not run with sample data)
results <- sim_schools(
    origins = origins_almada,
    destinations = schools_lisbon,
    model = "gravity",
    max_dist = 3000,
    balancing = "destinations",
    d = distance_euclidean,
    m = origin_pupils,
    n = destination_n_pupils,
    constraint_production = origin_pupils,
    beta = 0.99944,
    keep_cols = TRUE
)
```

## 3. Route Generation

Generate cycling routes between origins and the selected school:

```{r}
#| label: routes
# The routes are already generated in our sample data
# Here's how you would generate them:
# routes_almada <- bici_routes(od_data_almada, distance.threshold = 5e3)

# Visualize the routes
tm_shape(origins_almada) +
    tm_borders(alpha = 0.3) +
    tm_shape(od_data_almada) +
    tm_lines(col = "grey", alpha = 0.3) +
    tm_shape(routes_almada) +
    tm_lines(col = "dodgerblue", alpha = 0.6, lwd = 2) +
    tm_shape(schools_lisbon) +
    tm_dots(col = "red", size = 0.2) +
    tm_layout(title = "Cycling Routes to School")
```

## 4. Network Analysis and Cycling Uptake

Calculate potential cycling uptake using PCT models and create a route network:

```{r}
#| label: uptake-network
# Calculate cycling uptake potential
routes_uptake <- routes_pct_uptake(routes_almada)

# Create route network
rnet <- routes_network(routes_uptake)

# Summarize routes by origin
routes_summaries <- summarise_routes(routes_uptake)

cat("Routes with cycling potential:", nrow(routes_uptake), "\n")
cat("Network segments:", nrow(rnet), "\n")
```

Visualize the network with cycling potential:

```{r}
#| label: network-map
tm_shape(rnet) +
    tm_lines(
        col = "bicycle_godutch",
        palette = "viridis",
        lwd = 3,
        title.col = "Potential Cyclists"
    ) +
    tm_shape(schools_lisbon) +
    tm_dots(col = "red", size = 0.3) +
    tm_layout(title = "Cycling Network with Uptake Potential")
```

## 5. Bike Bus Route Identification

Identify optimal bike bus routes using the route bundling algorithm:

```{r}
#| label: bike-bus-routes
# Generate ordered bike bus routes
ordered_routes <- cycle_bus_routes(
    routes_summaries,
    rnet = rnet,
    min_trips = 1,
    buffer = 10
)

# Calculate route statistics
route_stats <- calc_stats(
    routes = routes_summaries,
    rnet_plan = rnet,
    ordered_routes = ordered_routes,
    min_trips = 1
)

# Filter to top routes
top_routes <- filter_routes(ordered_routes, buffer = 300, top_n = 3)

cat("Total potential routes:", nrow(ordered_routes), "\n")
cat("Top priority routes:", nrow(top_routes), "\n")
```

## 6. Centroid Matching and Final Selection

Match student origins to the top bike bus routes:

```{r}
#| label: centroid-matching
# Match centroids to bike bus routes
centroids <- match_centroids(
    routes_cents = ordered_routes,
    top_routes = top_routes,
    route_stats = route_stats,
    origins = origins_almada,
    attribute_trips = "bicycle_godutch"
)

cat("Students assigned to bike bus routes:", nrow(centroids), "\n")
```

## 7. Final Visualization

Create a comprehensive map showing the final bike bus network:

```{r}
#| label: final-map
# Create final visualization
tm_shape(origins_almada) +
    tm_borders(alpha = 0.2) +
    tm_shape(centroids) +
    tm_bubbles(
        col = "pick",
        size = "bicycle_godutch",
        palette = "Set1",
        title.col = "Bike Bus Route",
        title.size = "Potential Cyclists"
    ) +
    tm_shape(top_routes %>% mutate(route_id = as.character(row_number()))) +
    tm_lines(
        col = "route_id",
        palette = "Set1",
        lwd = 4,
        title.col = "Bike Bus Routes"
    ) +
    tm_shape(schools_lisbon) +
    tm_dots(col = "red", size = 0.5, shape = 17) +
    tm_layout(title = "Optimized Bike Bus Routes for School Travel")
```


